Relative parabolic factorizations of Chevalley groups

### 2015-04-21 stability-conditions.tex
Rewritten stable ranks section

As we will be mainly concerned with applications to stability of $K_1$ of Chevalley groups, our treatment of stability conditions is restricted to the case of commutative $R$. 
We refer the reader to~\cite{Ba64, Va69, Va71} for a more general exposition.

\subsection{Stable rank}
Recall the definition of \emph{stable rank} introduced by H.~Bass and L.~Vaserstein (see~\cite{Ba64, Va69}).
\begin{dfn}
A row $a=(a_1,\ldots,a_n)\in{}^n\!R$ is called \emph{$I$-unimodular} if it equals $(1,0,\ldots,0)$ modulo $I$ and $a_1,a_2,\ldots,a_n$ generate $R$ as an ideal.
\end{dfn}
A column $b \in R^n$ will be called $I$-unimodular if its transpose $b^t$ is an $I$-unimodular row. We denote the set of all $I$-unimodular rows (resp. columns) by $\Umd(n,R,I)$ (resp. $\Ums(n,R,I)$).
When $R=I$ we refer to $R$-unimodular rows and columns as unimodular.
\begin{lemma}\label{lemma:relstrlemma}
For any $I$-unimodular row $a=(a_1,\ldots,a_n)$ there exists $I$-unimodular column $b=(b_1,\ldots,b_n)^t$ such that $ab=\sum\limits_{i=1}^n a_i b_i=1$.
\end{lemma}
\begin {dfn} An $I$-unimodular row $a=(a_1,\ldots a_{n+1})$ is called \emph{stable} if one can choose $b_1,\ldots,b_n\in R$ such that the row $(a_1+a_{n+1}b_1,\ldots,a_n+a_{n+1}b_n)$ is also $I$-unimodular. If, moreover, elements $b_1,\ldots,b_n$ can be chosen from  $I$ then such $a$ will be called \emph{$I$-stable}. \end{dfn} 
We say that a pair $(R,I)$ satisfies stable range condition $\SR_n(R,I)$ if any $I$-unimodular row of length $n+1$ is stable.
\begin{lemma}\label{lemma:relstrlemma2}\strut
\begin{enumerate}
\item The condition $\SR_n(R,I)$ implies $\SR_m(R,I)$ for any $m\geqslant n$.
\item The condition $\SR_n(R,I)$ implies that any $I$-unimodular row of length $n+1$ is $I$-stable.
\end{enumerate}
\end{lemma}
By definition, the {\it relative stable rank} $\sr(R,I)$ of a pair $(R,I)$ is the smallest natural number $n$ such that condition $\SR_n(R,I)$ holds.
If no such $n$ exists, we set $\sr(R,I)=\infty$.

The following proposition summarizes basic properties of stable ranks.
\begin{prop} \label{prop:sr_properties}
Let $R$ be arbitrary commutative unital ring and let $I\trianglelefteq R$ be an ideal.
\begin{enumerate}
\item For any ideal $J\subseteq I$ one has $\sr(R,I)\leqslant\sr(R)$, $\sr(R/J,I/J)\leqslant\sr(R,I)$.
\item One has $\sr\left(\prod\limits_{i=1}^n R_i\right)=\max\limits_{i=1}^n\left(\sr(R_i)\right).$
\item Let $J$ denote the Jacobson radical of $R$, then one has $\sr(R)=\sr(R/J)$ and $\sr(R, J) = 1$.
\end{enumerate}
\end{prop}
\begin{proof} See~\cite{Va71}. 
\textbf{TODO: Add reference for the latter fact}.
%TODO:
\end{proof}
\begin{example}
Since the stable rank of a field equals $1$, one can conclude from the previous proposition that $\sr(R)=1$ for any semilocal ring $R$.
Other examples of rings of stable rank $1$ are provided by boolean rings, the ring of algebraic integers, the ring of entire functions, the disk-algebra.
\end{example}

@@

The next lemma illustrates a typical application of the absolute relative stability condition.
Let $v\in R^{2\ell}$ be a vector of the standard vector representation of $\G(\rD_\ell, R)$.
It will be convenient for us to denote by $v^+$ and $v^-$ the upper and lower halves of $v$ i.e. $v^+=(v_1,\ldots, v_\ell)^t$, $v^-=(v_{-\ell},\ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
 Let $\Phi=\rD_\ell$ and assume that $\asr(R, I)\leq \ell -1$. Then for any $I$-unimodular column $v=(v^+, v^-)^t\in\Ums(2\ell, R, I)$
 there exist $g\in\U(\Sigma^+_\ell, I)$ such that $(v\cdot g)^+ \in \Ums(\ell, R, I)$.
\end{lemma}
\begin{proof} Denote by $J$ the ideal of $R$ spanned by entries of $v^-$. Clearly, $J \leq I$.
By Proposition~\ref{prop:sr_properties} $\sr(I/J) \leq \ell-1$ and therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, R/J, I/J)$. 
This implies existence of $h\in \E(\Delta_\ell, R, I)$ such that for $v' = v\cdot h$ one has $(v')^+ \equiv e_1$ modulo $J$.

>>>>

As we will be mainly concerned with applications to stability of $K_1$ of Chevalley groups, our treatment of stability conditions is restricted to the case of commutative $R$
notwithstanding with the fact that the majority of constructions related to stable range conditions can be defined in a more general non-commutative setting (see e.g.~\cite{Ba64, Va69, Va71}).

\subsection{Stable rank}
Recall the definition of \emph{stable rank} introduced by H.~Bass and L.~Vaserstein (see~\cite{Ba64, Va69}).
\begin{dfn} A row $a\in{}^n\!R$ is called \emph{$I$-unimodular} if it is congruent to $e_1$ modulo $I$ and its components $a_1,a_2,\ldots,a_n$ generate $R$ as an ideal. \end{dfn}
A column $b \in R^n$ will be called $I$-unimodular if its transpose $b^t$ is an $I$-unimodular row.
We denote the set of all $I$-unimodular rows (resp. columns) by $\Umd(n,R,I)$ (resp. $\Ums(n,R,I)$).
When $R=I$ we refer to $R$-unimodular rows and columns as simply unimodular.

It is not hard to show that for a $I$-unimodular row $a$ there exists an $I$-unimodular column $b$ such that $ab=1$ (see~\cite[\S~2]{Va69}).

An $I$-unimodular row $a=(a_1,\ldots a_{n+1})$ is called \emph{stable} if one can choose $b_1,\ldots,b_n\in I$ such that the row $(a_1+a_{n+1}b_1,\ldots,a_n+a_{n+1}b_n)$ is also $I$-unimodular. 
We say that $I$ satisfies stable range condition $\SR_n(I)$ if any $I$-unimodular row of length $n+1$ is stable.
It is easy to see that $\SR_n(I)$ implies $\SR_m(I)$ for any $m\geqslant n$. It is also clear that validity of $\SR_n(I)$ does not depend on the ambient ring $R$.

By definition, the {\it stable rank} of $I$ is the smallest natural $n$ such that $\SR_n(I)$ holds.
We denote the stable rank of $I$ by $\sr(I)$, we set $\sr(I)=\infty$ if no such $n$ exists.

The following proposition summarizes basic properties of stable ranks.
\begin{prop} \label{prop:sr_properties}
Let $R$ be arbitrary commutative unital ring and let $I\trianglelefteq R$ be an ideal.
\begin{enumerate}
\item For any ideal $J\subseteq I$ one has $\sr(R,I)\leqslant\sr(R)$, $\sr(R/J,I/J)\leqslant\sr(R,I)$.
\item One has $\sr\left(\prod\limits_{i=1}^n R_i\right)=\max\limits_{i=1}^n\left(\sr(R_i)\right).$
\item Let $J$ denote the Jacobson radical of $R$, then one has $\sr(R)=\sr(R/J)$ and $\sr(R, J) = 1$.
\end{enumerate}
\end{prop}
\begin{proof} See~\cite{Va71}, \cite[Theorem~2.3]{Va69}. 
\textbf{TODO: Add reference for the latter fact}.
%TODO:
\end{proof}
\begin{example}
Since the stable rank of a field equals $1$, one can conclude from the previous proposition that $\sr(R)=1$ for any semilocal ring $R$.
Other examples of rings of stable rank $1$ are provided by boolean rings, the ring of algebraic integers, the ring of entire functions, the disk-algebra.
\end{example}

@@

The next lemma illustrates a typical application of the absolute relative stability condition.
Let $v\in R^{2\ell}$ be a vector of the standard vector representation of $\G(\rD_\ell, R)$.
It will be convenient for us to denote by $v^+$ and $v^-$ the upper and lower halves of $v$ i.e. $v^+=(v_1,\ldots, v_\ell)^t$, $v^-=(v_{-\ell},\ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
 Let $\Phi=\rD_\ell$ and assume that $\asr(R, I)\leq \ell -1$. Then for any $I$-unimodular column $v=(v^+, v^-)^t\in\Ums(2\ell, R, I)$
 there exist $g\in\U(\Sigma^+_\ell, I)$ such that $(v\cdot g)^+ \in \Ums(\ell, R, I)$.
\end{lemma}
\begin{proof} Denote by $J$ the ideal of $R$ spanned by entries of $v^-$. Clearly, $J \leq I$.
By Proposition~\ref{prop:sr_properties} $\sr(I/J) \leq \ell-1$ and therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, R/J, I/J)$ (see~\cite[Theorem~2.3c]{Va69}). 
This implies existence of $h\in \E(\Delta_\ell, R, I)$ such that for $v' = v\cdot h$ one has $(v')^+ \equiv e_1$ modulo $J$.
